Material Methods

### 2020-04-18 doc/MatMeth.Rmd
more edits

<This approach maintain breeding pairing and still result in prophase oocytes <(cite timeline of oocytes in neonates)>

@@

Meiocyte spreads were incubated with antibody protocol /  Immunohistochemistry was based on that in [@anderson1999] and [@koehler2002].

To visualize chromosomes/chromatin in the meiocytes -- we used markers for the centromere (CREST) and lateral element of of the SC (SYCP3). Double strand breaks (DSB) and COs were visualized with DMC1 and MLH1 respectively.

Antibodies and slide blocking was performed in 1X antibody dilution buffer (ADB) (normal donkey serum (Jackson ImmnuoResearch), 1X PBS, bovine serum albumin (Sigma), Triton X-100 (Sigma) ).

@@

Images of cells were capture using a Zeiss Axioplan 2 microscope with AxioLab camera and Axio Vision software (Zeiss, Cambridge, UK). Preprocessing, including cropping, noise reduction, and histogram adjustments, was performed using Photoshop (v13.0). Image file names were anonymized before manual scoring of MLH1 or DMC1 foci using photoshop.

@@

### DMC1
```{r DMC1.section.outline, echo=FALSE, include=FALSE}
#cell certeria, number, t.test
#DMC1 -- leptotene and sygotene cells were identified by SC morphology (by degree of synapsis) (cite another paper?). foci counted manually, error ccalulated from a subset of cells counted twice (supplement)

#<A subset of strains, *musculus^PWD^*, *musculus^MSM^*, *domesticus^WSB^*, and *domesticus^G^* were quantified for DMC1.  A total of `r length( (DMC1.DF %>% filter(stage == "Z"))$Original.Name)` and `r length( (DMC1.DF %>% filter(stage == "L"))$Original.Name)` of zygotene and leptotene cells were quantified respectively.>
```

<certeria for DMC1 stained cells (single cells) SC morphology used for staging.>

SC morphology was used to stage spermatocytes (as zygotene, leptotene, or pachtyene) (following X source). The means of **DMC1** foci per cell from a single mouse from a subset of srains *musculus^PWD^*, *musculus^MSM^*, *domesticus^WSB^*, and *domesticus^G^* were tested by t-tests.

@@

For **MLH1** pachytene quantification, cells with a full karyotype (19 acrocentric bivalents and XY for spermatocytes or 20 acrocentric bivalents for oocytes), distinct foci, and intact bivalents were included for quantification. (Quality score 1 to 5 )  based on foci brightness /staining and spreading of bivalents. The distributions of MLH1 counts per cell was assessed for normality (supplemental figure).

@@

The use of wild derived inbred strains enabled us to design an utilize a mixed model for our analytical framework. We built our model to predict the strain average MLH1 count per cell. The main effects and their interaction of subspecies and sex were coded as fixed. While, strain (or genetic background) was coded as a random effect to reflect a random sample of natural genetic variation from each subspecies. A strains to have a specific sex effect was also included.

We applied the following models to analze several dependant variables: mouse average mlh1 count per cell, the mouse average of variance in crossover count per cell, mouse mean SC length, mouse mean IFD^norm^ and IFD^raw^, and X.

@@

We chose to use a **mixed model framework** to interpret the patterns of variation, built using the lmer4 package (cite) in R (v3.5.2)[@Rstudio]. <We used the lme package - for estimating random variabkes>
<We used x package for testing significance of fixed effects and V package for testing significance of (strain) random variables (strain)>

@@

<to quantify / charaecterize the SC (length) (synapsed chromosomes, )-- two image analysis algorithms were used both which apply a 'skeletonizing' transformation to the synapsed chromosomes,  which transforms the elongatde chromosome space into a single pixel wide 'trace' of the originial chromosom shape.>

To quantify the SC area of indiviual cells and single chromosomes we applied the approaches discribed in [@wang_2019_sc] and [@peterson2019] respectively.  Both which apply a 'skeletonizing' transformation to the synapsed chromosomes,  which transforms the elongatde chromosome space into a single pixel wide 'trace' of the originial chromosom shape. Hand measures of SC/pactyene / meiotic chromosomes generated by using ImageJ/Fiji (v1.52) [@fiji].

##### total SC - Wang

As an additional metric of chromatin compaction, we computed the total (summed) SC length of all bivalents for single cells, using the image analysis algorithm of [@wang_2019_sc]. Mouse means were calculated from cell-wide total SC lengths in 2,984 -- `r length(  (MLH1.merge.Skel %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` out of 3,680 -- `r length(  (MLH1_data %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` cells with MLH1 counts (Figure X). the mouse average total sc was applied to the model framework.  <how else was this analyzed?!>

@@

The DNACrossover algorithm is hosted on CyVerse, more detailed description in @peterson2019. The DNA CrossOver algorithm isolated single chromosomes shapes from an image, in addition to returning the SC length and also measured the location of (green and blue signal along the single SC - reflecting MLH1 foci and cetromere signal respectively).

This algorithm substantially speeds the accurate measurement of bivalents, but has the limitation that not all bivalents per cell can be isolated due to overlapping bivalents. In this dataset, isolation rates per cell range from `r min(algorithm.stat.category$mean_pass.rate)` -- 0.51 (molossinus^MSM^ male) to `r max(algorithm.stat.category$mean_pass.rate)`-- 0.72 (musculus^KAZ^ female).

**curation step**  From the total set of cell images, `r length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName )` -- 10,458 bivalent objects were isolated by the image analysis software. After a human curation step ([@peterson2019]), `r length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID )` -- 9,829 single-bivalent observations remained.

To better understand the whole genome rec landscap -- we charecterizaed the number of crossover/MLH1 foci on single chromosomes (from the curated bivalent dataset).

We defined/classified each bivalent into a 'class' based on the number of MLH1 foci (range X to X). The **chromosome class proportions** different chromosome classes per strains by sex were tested with chi-sqaure tests with the prop.test function.

>>>>

@@

The immunohistochemistry was based on that in [@anderson1999] and [@koehler2002]. To visualize chromosomes in the meiocytes we used markers for the centromere (CREST) and lateral element of of the SC (SYCP3). Double strand breaks (DSB) and COs were visualized with DMC1 and MLH1 respectively. Antibodies and slide blocking was performed in 1X antibody dilution buffer (ADB) (normal donkey serum (Jackson ImmnuoResearch), 1X PBS, bovine serum albumin (Sigma), Triton X-100 (Sigma) ).

@@

Images were capture using a Zeiss Axioplan 2 microscope with AxioLab camera and Axio Vision software (Zeiss, Cambridge, UK). Preprocessing, including cropping, noise reduction, and histogram adjustments, was performed using Photoshop (v13.0). Image file names were anonymized before manual scoring of MLH1 or DMC1 foci using photoshop.

@@

### DMC1

```{r DMC1.section.outline, echo=FALSE, include=FALSE}
#cell certeria, number, t.test
#DMC1 -- leptotene and sygotene cells were identified by SC morphology (by degree of synapsis) (cite another paper?). foci counted manually, error ccalulated from a subset of cells counted twice (supplement)
```

SC morphology was used to stage spermatocytes in the zygotene, leptotene, or pachtyene cell stage. Mean **DMC1** foci per cell for a single mouse from a subset of srains *musculus^PWD^*, *musculus^MSM^*, *domesticus^WSB^*, and *domesticus^G^* were compared by t-tests.

@@

MLH1 foci were quantified from cells with intact and compelete karyotypes (19 acrocentric bivalents and XY for spermatocytes or 20 acrocentric bivalents for oocytes) and distinct MLH1 foci. A quality score, between 1 to 5, was asigned to each cell based on staining quality and general spread of bivalents. <cell with small defects were included -- and coded as a score of 5 - but these cell weren't included in the final analysis.>  The distributions of MLH1 counts per cell was assessed for normality (supplemental figure).

@@

The use of wild derived inbred strains enabled us to design an utilize a mixed model for our analytical framework. We built our model to predict the strain average MLH1 count per cell.

The main effects and their interaction of subspecies and sex were coded as fixed. While, strain (or genetic background) was coded as a random effect to reflect a random sample of natural genetic variation from each subspecies. A strains to have a specific sex effect was also included.

We applied the following models to analyze several dependent variables: mouse average mlh1 count per cell, the mouse average of variance in crossover count per cell, mouse mean total SC per cell, mouse mean single SC length, mouse mean IFD^norm^ and IFD^raw^, and position of crossover for single crossover bivalents.

X package used to test the fixed effects, Z package used to test the random effects.

@@

We chose to use a **mixed model framework** to interpret the patterns of variation, built using the lmer4 package (cite) in R (v3.5.2)[@Rstudio].

We used x package for testing significance of fixed effects and V package for testing significance of (strain) random variables (strain).

@@

To charaecterize the SC length of individual chromosomes which is a marker of chromatin compaction <from MLH1 stained cells>, two image analysis algorithms were used to specifically (measure indiviual cells) [@wang_2019_sc] and indiviaul chromosomes [@peterson2019]. Both apply a 'skeletonizing' transformation to the synapsed chromosomes which transforms the chromosome space into a single pixel wide 'trace' of the originial chromosom shape. Hand measures of SC/pactyene chromosomes were performed by using ImageJ/Fiji (v1.52) [@fiji].

The **total SC** per cell (all bivalents) was quantified across single cell images. Mouse means were calculated from cell-wide total SC lengths in 2,984 -- `r length(  (MLH1.merge.Skel %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` out of 3,680 -- `r length(  (MLH1_data %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` cells with MLH1 counts (Figure X). the mouse average total sc was applied to the model framework.  <how else was this analyzed?!>

@@

The **DNA CrossOver** algorithm isolates single straightened bivalents/chromosomes shapes from an image, and returns SC length, and location of green and blue signal, reflecting MLH1 foci and cetromere signal respectively. This algorithm substantially speeds the accurate measurement of bivalents, but has the limitation that not all bivalents per cell can be isolated due to overlapping bivalents. In this dataset, isolation rates per cell range from `r min(algorithm.stat.category$mean_pass.rate)` -- 0.51 (molossinus^MSM^ male) to `r max(algorithm.stat.category$mean_pass.rate)`-- 0.72 (musculus^KAZ^ female).

A curation step is applied to the total single bivalent dataset to remove poor isolation/segmentation instances. From the total set of cell images, `r length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName )` -- 10,458 bivalent objects were isolated by the image analysis software. After a human curation step ([@peterson2019]), `r length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID )` -- 9,829 single-bivalent observations remained.

**This dataset is used to charaecterize single bivalent traits**

The number of MLH1 foci on single chromosomes was quantified from the curated bivalent dataset to better understand the total recombination landscape. The **proportions** of single and double crossover bivalents were tested across mouse categories by chi-sqaure tests with the prop.test function.
